Spatio-angular fluorescence microscopy I. Basic theory (Draft: \today)

### 2018-11-07 papers/basic-theory/paper/paper.tex
Improved discussion of aplanatic imaging. Added pupil function sections.

\title{Spatio-angular fluorescence microscopy I. Basic theory}

\author{Talon Chandler,\authormark{1,*} Author order TBD,\authormark{2,*} and Patrick La Rivi\`ere \authormark{1}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling a microscope imaging an ensemble of in-focus
  fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular dipole transfer function. We demonstrate our formalism by
  analyzing a single-view paraxial epi-fluorescence microscope without using the
  monopole or scalar approximations. We show that this microscope has an angular
  band limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with simple phantoms. Notably, we
  show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is recorded by paraxial epi-fluorescence microscopes. We
  discuss the implications of our analysis for all quantitative fluorescence
  microscopy studies and lay out a path towards a complete theory.
\end{abstract*}

@@

This work lies at the intersection of three classes of fluorescence microscopy:
(1) single molecule localization microscopy (SMLM), (2) spatial ensemble
fluorescence microscopy including widefield, confocal, and light-sheet
techniques, and (3) polarized fluorescence microscopy. We briefly review these
three classes and focus on their use of the monopole and scalar approximations.

The SMLM community has pioneered the use of rigorous electromagnetic models of
fluorescence microscopes \cite{backer2014, lieb2004} [cite Novotny]. When a
single molecule is fluorescing in the sample, the measured intensity pattern is
strongly dependent on the orientation of the emitting molecule. Backlund [cite]
and Lew [cite] \cite{backlund2014} have shown that ignoring the orientation of
fluorophores can bias position estimates, so the most accurate SMLM experiments
must jointly estimate the position and orientation of each molecule.

Meanwhile, most fluorescence microscopists image ensembles of fluorophores and
restore their images without considering the role that the monopole and scalar
approximations play in the restoration process. A typical fluorescence
microscopist is only interested in the spatial distribution of fluorophores, so
they reason that they can ignore the orientation of the emitters.

A smaller community of microscopists is interested in measuring the orientation
of ensembles of fluorophores \cite{mehta2016} [Forkey, Goldman, Moerner,
Oldenbourg]. These techniques typically use polarized illumination or polarized
detection to make multiple measurements of the same object. Current angular
restoration techniques use a model of the dipole excitation and emission
processes [Fourkas] to recover the orientation of fluorophores using pixel-wise
arithmetic, but these techniques do not perform any spatial restoration so they
do not use all of the available information. To our knowledge no work has been
done to model the complete spatio-angular response of fluorescence microscopes
to ensembles of oriented fluorophores.

We begin in Section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently analyze
spatio-angular microscopes. In Section \ref{sec:results} we use the mathematical
tools we developed in Section \ref{sec:theory} to analyze and simulate a
single-view paraxial epi-fluorescence microscope, and we demonstrate our model
with four simple phantoms. Finally, in Section \ref{sec:discussion} we discuss
the results and their broader implications.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to draw new conclusions about all fluorescence microscopes. In
Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping in between
object and data in two different bases. In Section \ref{sec:dipole} we will
extend the monopole imaging model to dipoles and examine the mapping in four
different bases.

@@

We can represent the object as a function that assigns a real number to each
point on a plane, so we identify object space as
$\mbb{U} = \mbb{L}_2(\mbb{R}^2)$. Similarly, we have a two-dimensional detector
that measures a real number at every point on a plane, so data space is the same
set $\mbb{V} = \mbb{L}_2(\mbb{R}^2)$.

@@

Next, we assume that the optical system is \textit{aplanatic}---Abbe's sine
condition is satisfied and on-axis points are imaged without aberration. Abbe's
sine condition guarantees that off-axis points are imaged without spherical
aberration or coma \cite{mansuripur2009}, so the imaging system can be modeled
(at least approximately) as a magnifier with shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro)f(\ro), \label{eq:nonconv}
\end{align}
where $m$ is a magnification factor.

@@

The mapping between the object and the data in linear shift-invariant
microscopes takes a particularly simple form in a spatial harmonic basis. If we
apply the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bv) = H(\bv)F(\bv),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bv) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bv) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv).
\end{align}
The monopole point spread function is normalized and real-valued, so we know
that the monopole transfer function is normalized $H(0) = 1$ and
conjugate-symmetric $H(-\bv) = H^*(\bv)$.

@@

We have used the term \textit{monopole transfer function} instead of the more
common \textit{optical transfer function}. We define the optical transfer
function as the mapping between an input irradiance spectrum and an output
irradiance spectrum in an optical system. With this definition the optical
transfer function is not necessarily the same as the monopole transfer function.
For example, consider a light sheet illumination scheme where a sheet of the
sample is illuminated and an orthogonal detection arm collects the emitted
fluorescence. Both the excitation path and the detection path can be described
by optical transfer functions, and the monopole transfer function is given by
the convolution of the two optical transfer functions. We will see that this is
a valuable distinction when we consider dipole transfer functions.

@@

Since the monopole transfer function is given by the two-dimensional Fourier
transform of the monopole point spread function
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\,h(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bv],
\end{align}
we can plug in Eq. \ref{eq:absquarescalar} and use the autocorrelation theorem to
rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\bvp\, C(\bvp)C^*(\bvp - \bv), 
\end{align}
where $z^*$ denotes the complex conjugate of $z$, and we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function
\begin{align}
  C(\bvp) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bvp].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
We have chosen the symbol $\bvp$ as the coordinate for the coherent
transfer function because the Fourier plane of the detector is often called the
\textit{pupil plane}.

The coherent transfer function provides a valuable shortcut for analyzing
microscopes since it is usually straightforward to calculate the field in the
aperture plane. A typical approach for calculating the transfer functions is to
(1) calculate the field in the aperture plane, (2) scale the field to find the
coherent monopole transfer function, then (3) use the relationships in Figure
\ref{fig:monopole-transfer-functions} to calculate the other transfer functions.

@@

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{dipole density}---the number of dipoles at
position $\ro{}$ per unit area oriented along $\so{}$ per unit solid angle.
Similar to the monopole case, we model the mapping between the object and the
irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so). 
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. If the imaging system is
linear and aplanatic we can write the integral transform as
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro, \so)f(\ro, \so). 
\end{align}

@@

\subsubsection{Angular dipole transfer function}
The spherical harmonics are another set of convenient basis functions. We can
change basis from spherical delta functions to spherical harmonics by applying
the generalized Plancharel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}\, p(\mh{s})q(\mh{s}) = \lmsum P_\ell^m Q_\ell^m, \label{eq:plan}
\end{align}
where $p(\mh{s})$ and $q(\mh{s})$ are arbitrary functions on the sphere, $P_\ell^m$ and $Q_\ell^m$ are their spherical Fourier transforms defined by
\begin{align}
  P_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, p(\mh{s})Y_\ell^{m*}(\mh{s}),
\end{align}
and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions defined in
Appendix \ref{sec:sph}. Eq. \ref{eq:plan} expresses the fact that scalar
products are invariant under a change of basis (see Eq. 3.78 of
\cite{barrett2004}). The left-hand side of Eq. \ref{eq:plan} is the scalar
product of $\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the
right-hand side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a
spherical harmonic function basis. Applying Eq. \ref{eq:plan} to Eq.
\ref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum H_\ell^m(\rd - \ro)F_\ell^m(\ro) \label{eq:atf-form}
\end{align}
where we have defined the \textit{angular dipole transfer function} as
\begin{align}
  H_\ell^m(\rd - \ro) = \int_{\mbb{S}^2}d\so\, h(\rd - \ro, \so)Y_{\ell}^{m*}(\so),\label{eq:atf-prep} 
\end{align}
and the \textit{angular dipole spectrum} as
\begin{align}
  F_\ell^m(\ro) = \int_{\mbb{S}^2}d\so\, f(\ro, \so)Y_{\ell}^{m*}(\so).
\end{align}
Since the dipole point spread function is normalized and real-valued, we know
that the angular dipole transfer function is normalized
$\int_{\mbb{R}^2}d\mb{r}\, H_0^0(\mb{r}) = 1$ and conjugate symmetric
$H_\ell^{-m}(\mb{r}) = (-1)^mH_\ell^{m*}(\mb{r})$.

@@

 \section{Results}\label{sec:results}
In this section we apply the tools we developed in the previous sections to
analyze a paraxial epi-fluorescence microscope. We start by reviewing the
derivation of the monopole point spread function in preparation for the
derivation of the orientation-dependent point spread function. We conclude the
section by analyzing the dipole imaging map in the four bases we described
earlier.

TODO: Define and discuss paraxial. 4f, illumination uniformity and ignoring in
monopole Where are excitation and detection dipoles parallel.

\subsection{Monopole point spread function of a paraxial epi-fluorescence microscope}
In this section we derive the form of the monopole point spread function
$h(\rd - \ro)$. First, we place a monopole emitter at the focal point of an
objective lens and calculate the field $U_{\text{bfp}}(\rp)$ created in the back
focal plane. We can model a paraxial thin lens as a quadratic phase element, so
it converts an incident spherical wave to a plane wave in the back focal plane
with a sharp cutoff at the exit pupil of the lens
\begin{align}
  U_{\text{bfp}}(\mb{r}_b) \propto \Pi\left(\frac{|\mb{r}_b|}{2\alpha}\right) \equiv
  \begin{cases}
    1\quad \text{if}\quad |\mb{r}_b| < \alpha,\\
    0\quad \text{else},
  \end{cases}
\end{align}
where $\rp$ is the two-dimensional coordinate in the back focal plane and
$\alpha$ is the collection half angle---see Figure XXX.

@@

TODO Expand. Make clear that later paper. We postpone a full discussion of SVD to find the object space singular functions
that span the measurement space of the instrument.

@@

\subsection{What determines the angular frequency cutoff?}
The spatial-frequency cutoff of a fluorescence microscope is well known to vary
as NA$/\lambda$---we can increase spatial resolution by increasing the NA of the
instrument or by choosing a fluorophore with a shorter emission wavelength.
Analogously, the angular-frequency spectrum of a fluorescence microscope depends
on both the instrument and the sample.

@@

\subsection{Characterizing real spatio-angular microscopes}
The theoretical model we present in this work is an extreme simplification of a
real microscope. We have ignored the effects of thick samples, optical
aberration, scattering, high-NA objectives, finite fluorescence lifetimes, and
interactions between fluorophores, among other. Because of this long list of
unknown effects, real experiments should make an attempt to characterize the
specific imaging system.

Typical fluorescence microscopy experiments start with the monopole
approximation and attempt to characterize the microscope by imaging
sub-diffraction beads. These images are assumed to be good approximations of the
monopole point spread function and this data can be used to restore (deconvolve)
the data taken in arbitrary experiments.

Imaging sub-diffraction beads is not enough to completely characterize the
response of a microscope to an oriented sample. We need at least three samples
with a known orientations to characterize the SATF, so isotropic beads are not
enough. Small beads may not be isotropic [cite Lew].

>>>>

\title{Spatio-angular fluorescence microscopy I. Basic theory (Draft: \today)}

\author{Talon Chandler,\authormark{1,*} Authors TBD,\authormark{2,*}Rudolf Oldenbourg\authormark{2,*}, and Patrick La Rivi\`ere \authormark{1}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular dipole transfer function. We demonstrate our formalism by
  analyzing a single-view fluorescence microscope without using the monopole or
  scalar approximations. We show that this microscope has an angular band limit,
  and we demonstrate the value of the transfer function approach by efficiently
  simulating the imaging process with simple phantoms. Notably, we show that
  information about the out-of-plane orientation of ensembles of in-focus
  fluorophores is recorded by fluorescence microscopes, and we show that these
  effects persist in the paraxial limit. We discuss the implications of our
  analysis for all quantitative fluorescence microscopy studies and lay out a
  path towards a complete theory.
\end{abstract*}

@@

This work lies at the intersection of three classes of fluorescence microscopy:
(1) ensemble imaging techniques where many fluorophores are imaged in parallel
and the goal is to find their concentration throughout the sample, (2) single
molecule localization microscopy (SMLM) techniques where fluorophores are sparse
in the sample and the goal is to find the position of each fluorophore, and (3)
polarized fluorescence techniques where many fluorophores are imaged in parallel
and the goal is to find their concentration and orientation. We briefly review
these three classes and focus on their use of the monopole and scalar
approximations.

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and many widely used restoration techniques make either or both of
the monopole and scalar approximations. The Gibson-Lanni model \cite{gibson89}
and the Born and Wolf model \cite{born1980} are two examples of scalar models
that are widely used to restore data collected by fluorescence microscopes, and
the Richards and Wolf model \cite{richards} is another widely-used model that
accounts for vectorial diffraction. All three of these models describe how
optical fields propagate through an optical system, but they are agnostic to the
object that is being imaged. In other words, these models are \textit{optical}
models, and when they are used to model fluorescent objects they implicitly make
the monopole approximation.

The SMLM community has pioneered the use of rigorous electromagnetic models for
dipole imaging systems that do not adopt the scalar or monopole approximations
\cite{backer2014, lieb2004, nov2006}. When a single molecule fluoresces, its
irradiance pattern on the detector is strongly dependent on the orientation of
the emitting molecule. Backlund et al. \cite{backlund2014} have shown have shown
that ignoring the orientation of fluorophores can bias position estimates, so
the most accurate SMLM experiments must jointly estimate the position and
orientation of each molecule. Some SMLM experiments use polarizers to estimate
the orientation of single molecules more accurately [Backer DNA paper, other
Moerner papers], and these works do not adopt the scalar or monopole
approximations.

A smaller community of microscopists is interested in measuring the orientation
and position of ensembles of fluorophores \cite{mehta2016} [Forkey, Goldman,
Moerner, Oldenbourg]. These techniques typically use polarized illumination or
polarized detection to make multiple measurements of the same object. Current
angular restoration techniques use a model of the dipole excitation and emission
processes [Fourkas] to recover the orientation of fluorophores using pixel-wise
arithmetic, but these techniques do not perform any spatial restoration so they
do not use all of the available information.

In this work we place these three communities on a common theoretical footing.
We begin in Section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently analyze
spatio-angular microscopes. In Section \ref{sec:results} we use the mathematical
tools we developed in Section \ref{sec:theory} to analyze and simulate a
single-view epi-fluorescence microscope, and we demonstrate our model with four
simple phantoms. Finally, in Section \ref{sec:discussion} we discuss the results
and their broader implications.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to draw new conclusions about all fluorescence microscopes. In
Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping between
object and data space in two different bases. In Section \ref{sec:dipole} we
will extend the monopole imaging model to dipoles and examine the mapping in
four different bases.

@@

We can represent the object as a function that assigns a real number to each
point on a plane, so we identify object space as
$\mbb{U} = \mbb{L}_2(\mbb{R}^2)$---the set of square-integrable functions on the
two-dimensional plane. Similarly, we have a two-dimensional detector that
measures a real number at every point on a plane, so data space is the same set
$\mbb{V} = \mbb{L}_2(\mbb{R}^2)$.

@@

Next, we assume that the optical system is \textit{aplanatic}---Abbe's sine
condition is satisfied and on-axis points are imaged without aberration. Abbe's
sine condition guarantees that off-axis points are imaged without spherical
aberration or coma \cite{mansuripur2009}, so the imaging system can be modeled
within the field-of-view of the optical system as a magnifier with
shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro)f(\ro), \label{eq:nonconv}
\end{align}
where $m$ is a magnification factor. Note that we have modeled the object and
data as continuous two-dimensional functions that extend to infinity. All real
objects must be restricted to a finite field-of-view and the data must be
discretely sampled, but we continue with continuous function with infinite
extent for convenience.

@@

The mapping between the object and the data in linear shift-invariant
microscopes takes a particularly simple form in a spatial harmonic basis. If we
apply the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bv) = H(\bv)F(\bv),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bv) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bv) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv).
\end{align}
The monopole point spread function is normalized and real-valued, so we know
that the monopole transfer function is normalized $H(0) = 1$ and
conjugate-symmetric $H(-\bv) = H^*(\bv)$ where $z^*$ denotes the complex
conjugate of $z$.

@@

Note that we have used the term \textit{monopole transfer function} instead of
the commonly-used term \textit{optical transfer function}. We reserve the term
\textit{optical transfer function} for optical systems---the optical transfer
function maps between an input irradiance spectrum and an output irradiance
spectrum in an optical system. We can use optical transfer functions to model
the optical fields in a microscope, but ultimately we are always interested in
measuring an object. Although the monopole transfer function can be identical to
the optical transfer function, we will find the distinction to be valuable when
we consider dipole transfer functions.

@@

Since the monopole transfer function is given by the two-dimensional Fourier
transform of the monopole point spread function
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\,h(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bv],
\end{align}
we can plug in Eq. \ref{eq:absquarescalar} and use the autocorrelation theorem to
rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\bvp\, C(\bvp)C^*(\bvp - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function
\begin{align}
  C(\bvp) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bvp].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
We have chosen the symbol $\bvp$ as the coordinate for the coherent
transfer function because the Fourier plane of the detector is often called the
\textit{pupil plane}.

The coherent transfer function provides a valuable shortcut for analyzing
microscopes since it is usually straightforward to calculate the field in the
pupil plane. A typical approach for calculating the transfer functions is to (1)
calculate the field in the pupil plane, (2) scale the field to find the
coherent monopole transfer function, then (3) use the relationships in Figure
\ref{fig:monopole-transfer-functions} to calculate the other transfer functions.

@@

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{dipole density}---the number of dipoles at
position $\ro{}$ per unit area oriented along $\so{}$ per unit solid angle.
Similar to the monopole case, we model the mapping between the object and the
irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so). 
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. If the optical system is aplanatic
we can write the integral transform as
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro, \so)f(\ro, \so). 
\end{align}

@@

\subsubsection{Angular dipole transfer function}
The spherical harmonics are another set of convenient basis functions. We can
change basis from spherical delta functions to spherical harmonics by applying
the generalized Plancharel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}\, p(\mh{s})q(\mh{s}) = \lmsum P_\ell^m Q_\ell^m, \label{eq:plan}
\end{align}
where $p(\mh{s})$ and $q(\mh{s})$ are arbitrary functions on the sphere, $P_\ell^m$ and $Q_\ell^m$ are their spherical Fourier transforms defined by
\begin{align}
  P_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, p(\mh{s})Y_\ell^{m*}(\mh{s}),
\end{align}
and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions defined in
Appendix \ref{sec:sph}. Eq. \ref{eq:plan} expresses the fact that scalar
products are invariant under a change of basis (see Eq. 3.78 of
\cite{barrett2004}). The left-hand side of Eq. \ref{eq:plan} is the scalar
product of $\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the
right-hand side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a
spherical harmonic function basis. Applying Eq. \ref{eq:plan} to Eq.
\ref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum H_\ell^m(\rd - \ro)F_\ell^m(\ro), \label{eq:atf-form}
\end{align}
where we have defined the \textit{angular dipole transfer function} as
\begin{align}
  H_\ell^m(\rd - \ro) = \int_{\mbb{S}^2}d\so\, h(\rd - \ro, \so)Y_{\ell}^{m*}(\so),\label{eq:atf-prep} 
\end{align}
and the \textit{angular dipole spectrum} as
\begin{align}
  F_\ell^m(\ro) = \int_{\mbb{S}^2}d\so\, f(\ro, \so)Y_{\ell}^{m*}(\so).
\end{align}
Since the dipole point spread function is normalized and real-valued, we know
that the angular dipole transfer function is normalized
$\int_{\mbb{R}^2}d\mb{r}\, H_0^0(\mb{r}) = 1$ and conjugate symmetric
$H_\ell^{-m}(\mb{r}) = (-1)^mH_\ell^{m*}(\mb{r})$.

@@

\section{Results}\label{sec:results}
In this section we begin to apply the tools we developed in the previous section
to analyze more specific fluorescence microscopes. During our initial discussion
we only required the optical system to be aplanatic so that we could model it as
a shift-invariant magnifier. Now we will be more specific and consider an
aplanatic optical system in a \textit{$4f$ configuration} with an arbitrary
first lens (the objective) and a \textit{paraxial second lens} (the tube lens).
A lens can be considered paraxial if the angle $\alpha$ between the optical axis
of the lens and the marginal ray is small enough that
$\sin\alpha \approx \alpha$. As a rule of thumb, paraxial effects become
significant when the numerical aperture of a lens exceeds 0.7 \cite{gu2000}, but
this is only a rough guideline.

We start by defining \textit{pupil functions} for both monopole and dipole
imaging systems, and we explicitly relate the pupil function to the previously
defined coherent transfer function. Next, we calculate the transfer functions
for monopole and dipole imaging systems with high-NA objectives. We resort to
numerics as necessary, but we can calculate several transfer function in closed
form. We compare these results with paraxial results which are completely in
closed form.

\subsection{Monpole pupil functions}
We define the \textit{monopole pupil function} $p(\rp)$ of the imaging system as
the field immediately following the aperture plane created by an on-axis
monopole where $\rp$ is an unscaled two-dimensional coordinate in the pupil
plane. In this section we will relate the monopole pupil function to the
transfer function we defined in Section \ref{sec:theory} by following a near
variant of the path in Barrett and Myers Section 9.7 \cite{barrett2004}.

Since monopoles emit scalar fields, the monpole pupil function is a
scalar-valued function. The optical system is aplanatic, so we can write the
field created at a point in the aperture $\rp$ by a monopole at position $\ro$
as
\begin{align}
   U_p(\rp, \ro) \propto p(\rp)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]. \label{eq:pupil}
\end{align}
Eq. \ref{eq:pupil} is a restatement of the aplanatic condition for a $4f$
optical system---the field in the aperture plane can be written as the pupil
function multiplied by a linear phase factor that encodes the position of the
object.

Since the second lens is paraxial, we can model with relationship between the
field in the aperture plane and the field on the detector with a scaled Fourier
transform \cite{goodman1996}
\begin{align}
   U_d(\rd', \ro) \propto \int_{\mbb{R}^2}d\rp\, p(\rp)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]\text{exp}\left[-2\pi i \frac{n_1}{\lambda f_1} \rp\cdot\rd' \right]. \label{eq:detectorfourier}
\end{align}
If we define $P(\bvp)$ as the two-dimensional Fourier transform of the pupil function
then we can rewrite Eq. \ref{eq:detectorfourier} as
\begin{align}
  U_d(\rd', \ro) \propto P\left(\frac{n_0}{\lambda f_0}\ro + \frac{n_1}{\lambda f_1}\rd'\right),
\end{align}
which we can simplify further in terms of $m = -\frac{f_1n_0}{f_0n_1}$
\begin{align}
  U_d(\rd' - m\ro) \propto P\left(\frac{n_1}{\lambda f_1}[\rd' - m\ro]\right),
\end{align}
The irradiance on the detector is given by the absolute square of the field so
\begin{align}
  h'(\rd' - m\ro) \propto \left|P\left(\frac{n_1}{\lambda f_1}[\rd' - m\ro]\right)\right|^2.
\end{align}
If we demagnify the coordinates $\rd = \rd'/m$ and the irradiance
$h(\rd - \ro) \propto h'(m[\rd - \ro])$ we find that the monopole point spread function
is related to the Fourier transform of the monopole pupil function by
\begin{align}
  h(\rd - \ro) \propto \left|P\left(-\frac{n_o}{\lambda f_o}[\rd - \ro]\right)\right|^2.
\end{align}
The monopole coherent spread function is the absolute square of the monopole point spread function so
\begin{align}
  c(\rd - \ro) \propto P\left(-\frac{n_o}{\lambda f_o}[\rd - \ro]\right).
\end{align}
The monopole coherent transfer function is the Fourier transform of the monopole coherent spread function so
\begin{align}
  C(\bvp) \propto p\left(\frac{\lambda f_o}{n_o}\bvp\right). \label{eq:monopupil}
\end{align}
Equation \ref{eq:monopupil} is the key result of this section---the monopole
coherent transfer function is a scaled monopole pupil function.

\subsection{Dipole pupil function} 
We define the \textit{dipole pupil function} $\mb{p}(\rp, \so)$ of the imaging
system as the electric field immediately following the aperture plane created by
a dipole oriented along $\so$. Since dipoles emit vector-valued electric fields,
the dipole pupil function is a vector-valued function. Almost all of the
arguments in the previous section carry over to the dipole case. Briefly, we can
write the electric field created at a point in the aperture $\rp$ by a dipole at
$\ro$ oriented along $\so$ as
 \begin{align}
   \mb{E}(\rp, \ro, \so) = \mb{p}(\rp, \so)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]. \label{eq:pupil2}
 \end{align}
 The second lens is paraxial, so we can find the field on the detector with a
 Fourier transform
 \begin{align}
   \mb{E}_d(\rd', \ro, \so) \propto \int_{\mbb{R}^2}d\rp\, \mb{p}(\rp, \so)\,\text{exp}\left[-2\pi i \frac{n_0}{\lambda f_0} \rp\cdot\ro \right]\text{exp}\left[-2\pi i \frac{n_1}{\lambda f_1} \rp\cdot\rd' \right]. \label{eq:detectorfourier2}
\end{align}
Note that the Fourier transform of a vector field is the Fourier transform of
its scalar-valued orthogonal components, so Eq. \ref{eq:detectorfourier2}
specifies two two-dimensional Fourier transforms. We follow the same
manipulations as the previous section and find that the dipole coherent transfer
function is a scaled dipole pupil function
\begin{align}
  \mb{C}(\bvp, \so) \propto \mb{p}\left(\frac{\lambda f_o}{n_o}\bvp, \so\right). \label{eq:monopupil}
\end{align}

\subsection{Monopole transfer functions}
Several works \cite{petrov2017, backlund2018} have modeled an aplanatic
fluorescence microscope imaging isotropic emitters with the scalar pupil
function
\begin{align}
  p(\rp) \propto (1 - r_p^2)^{-1/4}\Pi\left(\frac{r_p}{2f_o\sin\alpha}\right), 
\end{align}
where $\Pi(\cdot)$ is a rectangle function defined by
\begin{align}
  \Pi(x) = 
  \begin{cases}
    1\quad \text{if}\quad |x| < \frac{1}{2},\\
    0\quad \text{else},
  \end{cases}
\end{align}
which models the aperture stop of the objective, and $(1 - r_p^2)^{-1/4}$
conserves power before and after the objective. Applying Eq. XXX we see that the
coherent transfer function for this system is given by
\begin{align}
  C(\bvp)
\end{align}

In this section we calculate the form of the monopole point spread function
$h(\rd - \ro)$ when both lenses of the optical system are paraxial. First, we
place a monopole emitter at the focal point of the objective lens. We can model
a paraxial thin lens as a quadratic phase element, so it converts an incident
spherical wave to a plane wave in the back focal plane with a sharp cutoff at
the exit pupil of the lens
\begin{align}
  U_{\text{bfp}}(\mb{r}_b) \propto \Pi\left(\frac{|\mb{r}_b|}{2\alpha}\right) \equiv
  \begin{cases}
    1\quad \text{if}\quad |\mb{r}_b| < \alpha,\\
    0\quad \text{else},
  \end{cases}
\end{align}
where $\rp$ is the two-dimensional coordinate in the back focal plane and
$\alpha$ is the collection half angle---see Figure XXX.

@@

TODO Expand. Make clear that a later paper will include a full discussion of SVD
to find the object space singular functions that span the measurement space of
the instrument.

@@

TODO: Cite Tournier/Descoteux change of basis from SH to SM basis.

\subsection{What determines the angular frequency cutoff?}
The spatial-frequency cutoff of a fluorescence microscope is well known to be
2NA$/\lambda$---we can increase spatial resolution by increasing the NA of the
instrument or by choosing a fluorophore with a shorter emission wavelength.
Analogously, the angular-frequency spectrum of a fluorescence microscope depends
on both the instrument and the sample.

@@

\subsection{Towards more realistic models}
The theoretical model we present in this work is an extreme simplification of a
real microscope. We have ignored the effects of thick samples, refractive-index
mismatch, aberration, scattering, rotation of fluorophores, finite fluorescence
lifetimes, and interactions between fluorophores among others. Because of this
long list of unknown effects, real experiments will likely require extensions of
the models developed here.

TODO: Discuss/plan future papers. SVD?
Defocus/3D imaging:
$\mc{H}: \mbb{L}_2(\mbb{R}^3\times\mbb{S}^2) \rightarrow \mbb{L}_2(\mbb{R}^3)$.
Polarized illumination/detection imaging:
$\mc{H}: \mbb{L}_2(\mbb{R}^3\times\mbb{S}^2) \rightarrow \mbb{L}_2(\mbb{R}^3\times\mbb{S}^1)$.
Multiview polarized illumination imaging:
$\mc{H}: \mbb{L}_2(\mbb{R}^3\times\mbb{S}^2) \rightarrow \mbb{L}_2(\mbb{R}^3\times\mbb{S}^1)^V$.
